---
title: "Invacost database checks"
output:
  html_document:
    df_print: kable
    toc: yes
    toc_float: true
---

Load here the database file

```{r}
library(invacost)
library(readxl)
invacost <- as.data.frame(read_xlsx("d:/r/projects/invacost/data-raw/InvaCost_database_v4.1.xlsx",
                                    na = c("NA", "#N/A", "#DIV/0!", "#VALEUR!",
                                           "Unspecified", "Unknown",
                                           "unspecified"),
                                    guess_max = 10000))

# data(invacost)
```


# Check for incorrect inputs in numeric fields 
```{r}
invacost$Applicable_year <- as.numeric(invacost$Applicable_year)
invacost$Publication_year <- as.numeric(invacost$Publication_year)
invacost$Probable_starting_year <- as.numeric(invacost$Probable_starting_year)
invacost$Probable_ending_year <- as.numeric(invacost$Probable_ending_year)
invacost$Probable_starting_year_adjusted <- as.numeric(invacost$Probable_starting_year_adjusted)
invacost$Probable_ending_year_adjusted <- as.numeric(invacost$Probable_ending_year_adjusted)
```

If a message says that there are NAs, it means one of these numeric fields has an incorrect input

# Check taxonomy 
## Species
```{r}
invacost$unique.t.id <- do.call("paste", 
                                 invacost[, c("Kingdom", 
                                              "Phylum", 
                                              "Class", 
                                              "Order",
                                              "Family", 
                                              "Genus", 
                                              "Species")])

test_taxo <- invacost[-which(duplicated(invacost[, c("unique.t.id", "Species")])), ]

dupe_t <- test_taxo$Species[which(duplicated(test_taxo$Species))]
dupe_t <- dupe_t[-which(dupe_t %in% c(NA, "Diverse/Unspecified"))]


test_taxo <- test_taxo[test_taxo$Species %in% dupe_t, c("Cost_ID", "Kingdom", "Phylum", "Class", "Order",
                                                 "Family", "Genus", "Species")]

records_to_check <- test_taxo$Cost_ID[order(test_taxo$Species)]

# Errors to correct
test_taxo[order(test_taxo$Species), ]
records_to_check

```




## Genus
```{r}
invacost$unique.t.id <- do.call("paste", 
                                 invacost[, c("Kingdom", 
                                              "Phylum", 
                                              "Class", 
                                              "Order",
                                              "Family", 
                                              "Genus")])

test_taxo <- invacost[-which(duplicated(invacost[, c("unique.t.id", "Genus")])), ]

dupe_t <- test_taxo$Genus[which(duplicated(test_taxo$Genus))]
dupe_t <- dupe_t[-which(dupe_t %in% c(NA, "Diverse/Unspecified"))]


test_taxo <- test_taxo[test_taxo$Genus %in% dupe_t, c("Cost_ID", "Kingdom", "Phylum", "Class", "Order",
                                                 "Family", "Genus", "Species")]


records_to_check <- unique(c(records_to_check,
                             test_taxo$Cost_ID[order(test_taxo$Genus)]))

# Errors to correct
test_taxo[order(test_taxo$Genus), ]
records_to_check 
```


## Family
```{r}
invacost$unique.t.id <- do.call("paste", 
                                 invacost[, c("Kingdom", 
                                              "Phylum", 
                                              "Class", 
                                              "Order",
                                              "Family")])

test_taxo <- invacost[-which(duplicated(invacost[, c("unique.t.id", "Family")])), ]

dupe_t <- test_taxo$Family[which(duplicated(test_taxo$Family))]
dupe_t <- dupe_t[-which(dupe_t %in% c(NA, "Diverse/Unspecified"))]



test_taxo <- test_taxo[test_taxo$Family %in% dupe_t, c("Cost_ID", "Kingdom", "Phylum", "Class", "Order",
                                                 "Family", "Genus", "Species")]

records_to_check <- unique(c(records_to_check,
                             test_taxo$Cost_ID[order(test_taxo$Family)]))

# Errors to correct
test_taxo[order(test_taxo$Family), ]
records_to_check
```

## Order
```{r}
invacost$unique.t.id <- do.call("paste", 
                                 invacost[, c("Kingdom", 
                                              "Phylum", 
                                              "Class", 
                                              "Order")])

test_taxo <- invacost[-which(duplicated(invacost[, c("unique.t.id", "Family")])), ]

dupe_t <- test_taxo$Family[which(duplicated(test_taxo$Family))]
dupe_t <- dupe_t[-which(dupe_t %in% c(NA, "Diverse/Unspecified"))]



test_taxo <- test_taxo[test_taxo$Family %in% dupe_t, c("Cost_ID", "Kingdom", "Phylum", "Class", "Order",
                                                 "Family", "Genus", "Species")]

records_to_check <- unique(c(records_to_check,
                             test_taxo$Cost_ID[order(test_taxo$Family)]))

# Errors to correct
test_taxo[order(test_taxo$Family), ]
records_to_check
```


## Class
```{r}
invacost$unique.t.id <- do.call("paste", 
                                 invacost[, c("Kingdom", 
                                              "Phylum", 
                                              "Class")])

test_taxo <- invacost[-which(duplicated(invacost[, c("unique.t.id", "Class")])), ]

dupe_t <- test_taxo$Class[which(duplicated(test_taxo$Class))]
dupe_t <- dupe_t[-which(dupe_t %in% c(NA, "Diverse/Unspecified"))]



test_taxo <- test_taxo[test_taxo$Class %in% dupe_t, c("Cost_ID", "Kingdom", "Phylum", "Class", "Order",
                                                 "Family", "Genus", "Species")]


records_to_check <- unique(c(records_to_check,
                             test_taxo$Cost_ID[order(test_taxo$Class)]))

# Errors to correct
test_taxo[order(test_taxo$Class), ]
records_to_check 
```

## Phylum
```{r}
invacost$unique.t.id <- do.call("paste", 
                                 invacost[, c("Kingdom", 
                                              "Phylum")])

test_taxo <- invacost[-which(duplicated(invacost[, c("unique.t.id", "Phylum")])), ]

dupe_t <- test_taxo$Phylum[which(duplicated(test_taxo$Phylum))]
dupe_t <- dupe_t[-which(dupe_t %in% c(NA, "Diverse/Unspecified"))]



test_taxo <- test_taxo[test_taxo$Phylum %in% dupe_t, c("Cost_ID", "Kingdom", "Phylum", "Class", "Order",
                                                 "Family", "Genus", "Species")]


records_to_check <- unique(c(records_to_check,
                             test_taxo$Cost_ID[order(test_taxo$Phylum)]))

# Errors to correct
test_taxo[order(test_taxo$Phylum), ]
records_to_check 
```


## Comparison with GBIF taxonomy

The gbif check occurrs with the command line `taxize::classification()`. For a complete analysis,
the script should be run interactively, with the argument `ask = TRUE`, such that
in case of multiple matches in gbif, the user can decide which one to choose.

```{r}
invacost_sp <- unique(invacost$Species)
# Exclude multiple species and sp. / spp.
invacost_sp <- invacost_sp[-grep("/", invacost_sp)]
invacost_sp <- invacost_sp[-grep("sp\\.", invacost_sp)]
invacost_sp <- invacost_sp[-grep("spp\\.", invacost_sp)]
for(sp in invacost_sp)
{
  taxocheck <- unique(invacost[which(invacost$Species == sp), c("Kingdom",
                                                                "Phylum",
                                                                "Class",
                                                                "Order",
                                                                "Family", 
                                                                "Genus",
                                                                "Species")])
  
  # Next line is set to ask = FALSE for generating the Rmarkdown.
  # For a more detailed check, it should be set to ask = TRUE, and the analysis should be performed interactively
  taxo_correct <- try(taxize::classification(sp, db = "gbif", messages = FALSE, ask = TRUE))
  if (inherits(taxo_correct, "try-error"))
  {
    message("--> ", sp, " could not be resolved with the gbif database,
            check manually\n")
    next
  }
  
  if(all(is.na(taxo_correct[[1]])))
  {
    message("--> ", sp, " could not be resolved with the gbif database,
            check manually\n")
    next
  }
  
 if(nrow(taxocheck) > 1)
  {
    message("--> ", sp)
    message(paste0("Species ", sp, " has different taxonomies in invacost: \n\n"))
    message(paste0(do.call("paste", taxocheck), sep = "\n"))
    message("Correct taxonomy: ")
    message(paste0(apply(taxo_correct[[1]][, 2:1],
                                1, paste, collapse = " = ")
                          , sep = "\n "), collapse = " ")

    next
  }
  

  taxocheck <- rbind(taxocheck,
                     c(taxo_correct[[1]]$name[which(taxo_correct[[1]]$rank == "kingdom")],
                       taxo_correct[[1]]$name[which(taxo_correct[[1]]$rank == "phylum")],
                       taxo_correct[[1]]$name[which(taxo_correct[[1]]$rank == "class")],
                       taxo_correct[[1]]$name[which(taxo_correct[[1]]$rank == "order")],
                       taxo_correct[[1]]$name[which(taxo_correct[[1]]$rank == "family")],
                       taxo_correct[[1]]$name[which(taxo_correct[[1]]$rank == "genus")],
                       taxo_correct[[1]]$name[which(taxo_correct[[1]]$rank == "species")]
                       ))
  

  if(any(taxocheck[1, ] != taxocheck [2, ]) |
     is.na(any(taxocheck[1, ] != taxocheck [2, ])))
  {
    message("--> ", sp)
    message(paste0(" -  ",
                   colnames(taxocheck)[which(taxocheck[1, ] != taxocheck[2, ])],
                   ":\n              ***Invacost = ",
                   taxocheck[1, which(taxocheck[1, ] != taxocheck[2, ])],
                   " \n              ***Correct = ",
                   taxocheck[2, which(taxocheck[1, ] != taxocheck[2, ])],
                   sep = "\n"))
  }
}
taxocheck
```




# Geographic checks

## Country errors 
```{r}
if(!file.exists("c:/r/projects/invacost/data-raw/ne_50m_admin_0_countries.shp"))
{
  download.file("https://naciscdn.org/naturalearth/50m/cultural/ne_50m_admin_0_countries.zip",
                destfile = "data/tmp_data/ne_50m_admin_0_map_units.zip")
  unzip("data-raw/ne_50m_admin_0_map_units.zip",
        exdir = "data-raw")
  
  # countries <- rnaturalearth::ne_countries(scale = 50) # not up-to-date?
}

invisible(countries <- rgdal::readOGR("c:/r/projects/invacost/data-raw/ne_50m_admin_0_countries.shp"))



invacost.countries <- data.frame(original.name = unique(invacost$Official_country))

invacost.countries$valid <- FALSE
invacost.countries$valid[which(invacost.countries$original.name %in% countries$ADMIN)] <- TRUE

invacost.countries$renamed <- as.character(invacost.countries$original.name)
invacost.countries$renamed[grep("/", invacost.countries$original.name)] <- "Multiple countries"
invacost.countries$renamed[which(invacost.countries$original.name %in% c("African Countries", "European Union",
                                                                         "31 countries"))] <- "Multiple countries"

invacost.countries[!invacost.countries$valid & invacost.countries$renamed != "Multiple countries", ]
```

